---
title: "Regression Modelling of Car MPG vs. Transmission type"
author: "MattDSquared"
date: "Saturday, July 25, 2015"
output: html_document
---
```{r nonstandard.functions, echo=FALSE, results='hide', message=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
# credit: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
# author: winston@stdout.org
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

## Introduction
_from coursera.org class project description_
You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

"Is an automatic or manual transmission better for MPG"
"Quantify the MPG difference between automatic and manual transmissions"

R markdown code can be found at my github site: https://github.com/MattDSquared/RegressionModels-MotorCars/blob/master/motorcars.Rmd

## Setup and data cleaning

The mtcars dataset from 1981 will be used to answer these questions. This is located in the `datasets` library

```{r setup, echo=FALSE, message=FALSE}
library(datasets)
library(dplyr)
library(xtable)
# library(GGally)
library(ggplot2)

graphics.off()

setwd("~/../datascience/RegressionModels-MotorCars")
data(mtcars)

# ?mtcars

# rename vars
mtcars <- rename(mtcars, cylinders=cyl, displacement=disp, power=hp, 
                 axle.ratio=drat, weight=wt, quarter.mile.time=qsec, 
                 engine.shape=vs, transmission=am, n.gears=gear, 
                 n.carburetors=carb)

# move row names into it's own column
mtcars <-  mutate(mtcars, carnames=rownames(mtcars))
# move carnames to 1st column
mtcars <- mtcars[,c(ncol(mtcars),1:length(mtcars)-1)]
```

```{r cleaning, echo=c(1,2)}
# clarify values based on ?mtcars info
mtcars <- mutate(mtcars,
                 weight=weight*1000,
                 engine.shape=factor(engine.shape, levels=c(0,1), labels=c("v","straight")),
                 transmission=factor(transmission, levels=c(0,1),labels=c("automatic","manual")),
                 log.displacement=log(displacement),
                 log.weight=log(weight))

## for ease of using plot(myfit)
#rownames(mtcars) <- mtcars$carnames
# note: engine.shape has levels(0,1) = ("v","straight")
#       transmission has levels(0,1) = ("auto","manual")
```

```{r, echo=FALSE, results='asis'}
#tab <- xtable(head(mtcars))
#print(tab, type = "html", include.rownames=FALSE)
```

## Data Exploration

At first glance, it looks like manual is much better than automatic:

```{r first.glance, echo=FALSE, results='asis'}
mtcars.summary <- mtcars %>% group_by(transmission) %>% summarize(mpg.mean=mean(mpg))
tab <- xtable(mtcars.summary)
print(tab, type = "html", include.rownames=FALSE)
```

However, this does not take into account confounding factors, such as smaller cars tending to have manual drives. 

```{r weight.mpg.transmission, echo=FALSE, fig.height=4}
gg1 <- ggplot(mtcars, aes(x=log.weight, y=mpg, col=transmission)) +
    geom_point() + 
    geom_hline(yintercept=mtcars.summary$mpg.mean, size=1.5)+
    geom_smooth(method="lm") +
    geom_smooth(aes(x=log.weight, y=mpg), 
                method="lm", col="black", se=FALSE) +
    labs(title="MPG vs. Weight") + 
    theme(legend.position="top")

gg2 <- ggplot(mtcars, aes(x=log.displacement, y=mpg, 
                               col=transmission)) +
    geom_point() + 
    geom_hline(yintercept=mtcars.summary$mpg.mean, size=1.5)+
    geom_smooth(method="lm") +
    geom_smooth(aes(x=log.displacement, y=mpg), 
                method="lm", col="black", se=FALSE) +
    labs(title="MPG vs. Displacement and Transmission Type") + 
    theme(legend.position="top")

multiplot(gg1,gg2,cols=2)
```

## Model Selection
The ANOVA model search technique is used to identify variables of interest to our model. We begin with the effect from transmission type, since we are specifically interested in it, then observed the confidence interval for the changes in variance . A confidence of <.05 for significant change between models is designated as the model precision cuttoff. 

```{r ANOVA.model.search, echo=FALSE}
fit <- list()
fit[[1]] <- lm(mpg ~ transmission, data=mtcars)
fit[[2]] <- lm(mpg ~ transmission + log.weight, data=mtcars)
fit[[3]] <- lm(mpg ~ transmission + log.weight + log.displacement, data=mtcars)
fit[[4]] <- lm(mpg ~ transmission + log.weight + log.displacement +
               quarter.mile.time, data=mtcars)
fit[[5]] <- lm(mpg ~ transmission + 
                 log.weight + log.displacement +
                 cylinders + 
                 power + axle.ratio +
                 quarter.mile.time + engine.shape +
                 n.gears + n.carburetors, data=mtcars)

# for lack of more elegant way to do this, here's a string list of the models
model.name <- list()
for (item in sapply(fit, function(myfit) deparse(myfit$call))) {
    tmp2 <- ""
    for (entry in item) {
        tmp2 <- paste0(tmp2,entry)
    }
    model.name <- c(model.name,tmp2)
}
model.name <- unlist(gsub("     "," ", model.name))

# run ANOVA
fit.anova <- anova(fit[[1]], fit[[2]], fit[[3]], fit[[4]], fit[[5]])

# get the best model
bestmodel.num <- tail(which(fit.anova$"Pr(>F)" <= .05),n=1)
bestmodel <- fit[[bestmodel.num]]
bestmodel.str <- deparse(bestmodel$call)
bestmodel.str <- paste0(bestmodel.str[1],bestmodel.str[2])
```

```{r ANOVA.display, echo=FALSE, results='asis'}
tab <- xtable(data.frame(model.name))
print(tab, type = "html")
tab <- xtable(data.frame(fit.anova))
print(tab, type = "html")
```

The ANOVA results show that the model `r bestmodel.num`: `r bestmodel.str` will provide an optimal level of model accuracy without excess variance inflation.

## Model results on mpg vs. transmission type

```{r results.setup, echo=FALSE}
mtcars <- mutate(mtcars, 
                 mpg_hat = predict(bestmodel),
                 mpg_e = resid(bestmodel)) #mtcars$mpg - mpg_hat

# move mpg_hat and mpg_e to 3rd and 4th columns
mtcars <- mtcars[,c(1,2,ncol(mtcars)-1, ncol(mtcars), 3:(length(mtcars)-2))]
```

The residuals from this model show some variation near the ends of the data. However, for a linear model, this is a fairly acceptable fit, considering an R^2^ value of `r round(summary(bestmodel)$r.squared,3)` the model. 

```{r results.residuals, echo=FALSE, fig.height=4, fig.width=6.5}
gg1 <- ggplot(mtcars, aes(x=log.weight, y=mpg_e, col=transmission)) +
    geom_point() + 
    geom_smooth(aes(x=log.weight, y=mpg_e), col="black") +
    labs(title="Log(Weight) Residuals") + 
    theme(legend.position="top")
gg2 <- ggplot(mtcars, aes(x=log.displacement, y=mpg_e, 
                               col=transmission)) +
    geom_point() + 
    geom_smooth(aes(x=log.displacement, y=mpg_e), col="black") +
    labs(title="Log(displacement) Residuals") + 
    theme(legend.position="top")
multiplot(gg1,gg2,cols=2)
```

```{r results.confidence, echo=FALSE, results='asis'}
tab <- xtable(summary(bestmodel)$coeff)
print(tab, type = "html", include.rownames=FALSE)
```

### Model interpretation
The Intercept of `r round(summary(bestmodel)$coeff[1,1],1)` MPG indicates a not-so-useful value for when an automatic car has log(weight) and log(displacement) equal to 0. More importantly, for each 1 unit change in the log(weight), MPG is expected to decrease by `r round(summary(bestmodel)$coeff[3,1],1)`, holding all other variables constant. Similarly, for each 1 unit change in the log(displacement), MPG is expected to decrease by `r round(summary(bestmodel)$coeff[4,1],1)`, holding all other variables constant. Both of these rates have a significance level <= `r round(max(summary(bestmodel)$coeff[3:4,4]),3)`.

However, the effect of transmission on mpg is very small. In fact, the model shows a manual transmission will change mpg by `r round(summary(bestmodel)$coeff[2,1],1)`, but the standard error for this value is `r round(summary(bestmodel)$coeff[2,2],1)` and it has a p-value of `r round(summary(bestmodel)$coeff[2,4],3)`. 

## Conclusion

Given the high p-value of `r round(summary(bestmodel)$coeff[2,4],3)` and the small effect of `r round(summary(bestmodel)$coeff[2,1],1)` on mpg, no discernable difference in a car's fuel economy can be determined based on it's transmission type.

## Appendix

### Principle Component Analysis (PCA)

A brief PCA analysis shows highly redundant input data. 

```{r pca}
inputs <- select(mtcars, cylinders:n.carburetors)
mtcars.svd <- svd(scale(sapply(inputs, unclass)))
```

`r round(sum(mtcars.svd$d[1:2]^2/sum(mtcars.svd$d^2))*100,1)`% of the variation in input data can be explained by 2 variables. 

```{r pca.varexplain, echo=FALSE}
gg <- ggplot() + 
    geom_bar(aes(x=1:length(mtcars.svd$d), 
                 y=mtcars.svd$d^2/sum(mtcars.svd$d^2)),
             stat="identity",
             fill="dodgerblue") + 
    scale_x_discrete(limits=1:length(inputs)) +
    labs(title="Feature Variance") +
    labs(x="Orthogonal variables") +
    labs(y="Proportion of variance explained")
print(gg)
```

However, that variation would be a combination of nearly every input variable as shown below.

```{r pca.explore, echo=FALSE, warning=FALSE}
eigval <- 1
gg.svd.eigval <- ggplot() + 
    geom_bar(aes(x=1:length(inputs), y=mtcars.svd$v[,eigval]),
             stat="identity",
             fill="dodgerblue") + 
        scale_x_discrete(limits=names(inputs)) +
        labs(title=paste("Composition of eigenvalue",eigval)) +
        labs(x="Feature") +
        labs(y="Scaled Column Means")
print(gg.svd.eigval)
```

### Preliminary Plot of Mpg vs. features

```{r ggfunction, echo=FALSE, results='hide'}
# consider avoiding ggplot calls inside of a function. Behavior can be unpredictable
plot.mpg.by <- function(var, main=NULL) {
    ggplot(mtcars, aes_string(x=var, y="mpg", col="transmission")) +
        geom_point() + 
        geom_smooth(method="lm") +
        geom_smooth(aes_string(x=var, y="mpg"), method="lm", 
                    col="black", se=FALSE) +
        labs(title=main)
}
```

```{r exploration.rawdata, echo=FALSE, fig.width=8, fig.height=14}
# MPG vs feature broken out between automatic and manual
gg  <- list()
ii <- 0
for (feat.name in names(mtcars)[3:ncol(mtcars)]) {
    #tmp <- select(mtcars, mpg, transmission, feat.name)
    #rename(tmp, )
    ii <- ii + 1
    gg[[ii]] <- plot.mpg.by(feat.name)
}
#windows(width = 9, height = 10, xpos = 0, rescale="fixed")
multiplot(plotlist = gg, cols=2)

#windows(width = 13, height = 11, xpos = 30)
#pairs(select(mtcars, -carnames), panel = panel.smooth, main = "Motor Cars", 
#      col = 3 + (mtcars$log.displacement > 5.25))

# ggpairs(mtcars, lower=list(continuous = "smooth"), params=c(method="lm"))
```

### Post-analysis plots of Residual MPG vs. features

```{r exploration.residuals, echo=FALSE, fig.width=8, fig.height=14}
# MPG vs feature broken out between automatic and manual

plot.mpg_e.by <- function(var, main=NULL) {
    ggplot(mtcars, aes_string(x=var, y="mpg_e", col="transmission")) +
        geom_point() + 
        geom_smooth(method="lm") +
        geom_smooth(aes_string(x=var, y="mpg_e"), method="lm", 
                    col="black", se=FALSE) +
        labs(title=main)
}

gg  <- list()
ii <- 0
for (feat.name in names(mtcars)[3:ncol(mtcars)]) {
    #tmp <- select(mtcars, mpg, transmission, feat.name)
    #rename(tmp, )
    ii <- ii + 1
    gg[[ii]] <- plot.mpg_e.by(feat.name)
}
#windows(width = 9, height = 10, xpos = 0, rescale="fixed")
multiplot(plotlist = gg, cols=2)

#windows(width = 13, height = 11, xpos = 30)
#pairs(select(mtcars, -carnames), panel = panel.smooth, main = "Motor Cars", 
#      col = 3 + (mtcars$log.displacement > 5.25))

# ggpairs(mtcars, lower=list(continuous = "smooth"), params=c(method="lm"))
```

```{r exploration.pca.multiplot, echo=FALSE, fig.width=8, fig.height=10}
# This doesn't work because of ggplot's inability to calculate valiables from local scope
#gg  <- list()
#for (ii in 1:length(mtcars.svd$d)) {
#    #tmp <- select(mtcars, mpg, transmission, feat.name)
#    #rename(tmp, )
#    eigval <<- ii
#    gg[[ii]] <- ggplot(environment=environment()) + 
#        geom_bar(aes(x=1:length(inputs), y=mtcars.svd$v[,eigval]),
#                 stat="identity",
#                 fill="dodgerblue") + 
#        scale_x_discrete(limits=names(inputs)) +
#        labs(title=paste("Composition of eigenvalue",eigval)) +
#        labs(x="Feature") +
#        labs(y="Scaled Column Means")
#}
#windows(width = 9, height = 10, xpos = 0, rescale="fixed")
#multiplot(plotlist = gg, cols=2)
```