---
title: "Regression Modelling of Car MPG vs. Transmission type"
author: "MattDSquared"
date: "Saturday, July 25, 2015"
output: html_document
---

```{r nonstandard.functions, echo=FALSE, results='hide', message=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

## Introduction
_from coursera.org class project description_
You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

"Is an automatic or manual transmission better for MPG"
"Quantify the MPG difference between automatic and manual transmissions"

R markdown code can be found at my github site: https://github.com/MattDSquared/RegressionModels-MotorCars/blob/master/motorcars.Rmd

## Setup and data cleaning

The mtcars dataset from 1981 will be used to answer these questions. 

```{r setup, echo=FALSE, message=FALSE}
library(datasets)
library(dplyr)
library(xtable)
library(ggplot2)

setwd("~/../datascience/RegressionModels-MotorCars")
data(mtcars)

# rename vars
mtcars <- rename(mtcars, cylinders=cyl, displacement=disp, power=hp, 
                 axle.ratio=drat, weight=wt, quarter.mile.time=qsec, 
                 engine.shape=vs, transmission=am, n.gears=gear, 
                 n.carburetors=carb)
```

```{r cleaning, results='asis'}
# clarify values based on ?mtcars info
mtcars <- mutate(mtcars,
                 weight=weight*1000,
                 engine.shape=factor(engine.shape, levels=c(0,1), 
                                     labels=c("v","straight")),
                 transmission=factor(transmission, levels=c(0,1),
                                     labels=c("automatic","manual")))
```

```{r, echo=FALSE, results='asis'}
tab <- xtable(head(mtcars))
print(tab, type = "html")
```

## Data Exploration

```{r exploration.plots, echo=FALSE, fig.width=8, fig.height=14}
# MPG vs feature broken out between automatic and manual
gg  <- list()

ii <- 0
for (feat.name in names(mtcars)[2:ncol(mtcars)]) {
    #tmp <- select(mtcars, mpg, transmission, feat.name)
    #rename(tmp, )
    ii <- ii + 1
    gg[[ii]] <- ggplot(mtcars, aes_string(x=feat.name, 
                                          y="mpg", 
                                          col="transmission")) +
        geom_point() + 
        geom_smooth(method="lm") 
}

multiplot(plotlist = gg, cols=2)
```