---
title: "Regression Modelling of Car MPG vs. Transmission type"
author: "MattDSquared"
date: "Saturday, July 25, 2015"
output: html_document
---

## Introduction
_from coursera.org class project description_
You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

"Is an automatic or manual transmission better for MPG"
"Quantify the MPG difference between automatic and manual transmissions"

## Setup and data cleaning

The mtcars dataset from 1981 will be used to answer these questions. 

```{r setup, echo=FALSE, message=FALSE}
library(datasets)
library(dplyr)
library(xtable)
library(ggplot2)

setwd("~/../datascience/RegressionModels-MotorCars")
data(mtcars)
```

```{r cleaning, results='asis'}
# rename vars
mtcars <- rename(mtcars, cylinders=cyl, displacement=disp, power=hp, 
                 axle.ratio=drat, weight=wt, quarter.mile.time=qsec, 
                 engine.shape=vs, transmission=am, n.gears=gear, 
                 n.carburetors=carb)

# clarify values based on ?mtcars info
mtcars <- mutate(mtcars,
                 weight=weight*1000,
                 engine.shape=factor(engine.shape, levels=c(0,1), 
                                     labels=c("v","straight")),
                 transmission=factor(transmission, levels=c(0,1),
                                     labels=c("automatic","manual")))
```

```{r, echo=FALSE, results='asis'}
tab <- xtable(head(mtcars))
print(tab, type = "html")
```

## Data Exploration

